---
title: "ps05-rmarkdown-plot"
author: "Yixiang Fang"
date: "2023-02-20"
output: html_document
---

```{r message = FALSE}
library(tidyverse)
```

# 1 Load and check data 
### Q2
```{r message = FALSE}
gpminder <- read_delim("data/gapminder.csv.bz2")
```

```{r}
dim(gpminder)
```
13055 rows and 25 columns

### Q3
```{r}
gpminder %>%
  sample_n(10)
```
There are many N/A values in the dataset

# 2 Descriptive statistics 
### Q1
```{r}
num_iso3 <- length(unique(gpminder$iso3))
num_iso3
num_iso2 <- length(unique(gpminder$iso2))
num_iso2
num_name <- length(unique(gpminder$name))
num_name
```
There are 253 countries according to iso3. There are 249 countries according to iso2. There are 250 countries according to name.

### Q2
### a
```{r}
gpminder %>% 
  group_by(iso2) %>% 
  summarize(n = n_distinct(name)) %>%
  filter(n > 1)
```

```{r}
cor_countries <- unique(gpminder$name[is.na(gpminder$iso2)])
cor_countries
```
There are iso2 code correspond to one name. The country is Namibia when the iso2 is NA.

### b
```{r}
gpminder %>% 
  group_by(name) %>% 
  summarize(n = n_distinct(iso3)) %>%
  filter(n > 1)
```

```{r}
more_iso3 <- unique(gpminder[is.na(gpminder$name), ]$iso3)
more_iso3
```
It has 4 iso3-code when the name is mission. Four countries are Chanisl, GBM, KOS, and NLD_CURACAO.

### Q3
```{r}
min(gpminder$time, na.rm = TRUE)
max(gpminder$time, na.rm = TRUE)
```

The minimum year is 1960, and the maximum year is 2019.

# 3 CO2 emissions
### Q1
```{r}
gpminder %>% 
  group_by(time) %>% 
  summarize(n = sum(is.na(co2))) %>%
  arrange(desc(n)) %>% 
  head()
```

```{r}
gpminder %>% 
  group_by(time) %>% 
  summarize(n = sum(is.na(co2_PC))) %>%
  arrange(desc(n)) %>% 
  head()
```
2017, 2018, 2019 have the largest value of missing co2 emission.

### Q2
```{r}
gpminder %>% 
  filter(name %in% c("United States of America", "China", "India", "Argentina", "Zimbabwe")) %>%
  ggplot(aes(x = time, y = co2, color = name)) + geom_line()
```
For China, India, and US are increasing for co2 emissions, but Argentina, and Zimbabwe are pretty flat.

### Q3
```{r}
gpminder %>% 
  filter(name %in% c("United States of America", "China", "India", "Argentina", "Zimbabwe")) %>%
  ggplot(aes(x = time, y = co2_PC, color = name)) + geom_line()
```
China, India, and Argentina are increasing, but the US are decreasing after 1980, same as Zimbabwe.

### Q4
```{r}
gpminder %>% 
  group_by(time, region) %>% 
  summarize(co2_pc = mean(co2_PC, na.rm = TRUE)) %>%
  na.omit() %>%
  ggplot(aes(x = time, y = co2_pc, color=region)) + geom_line()
```

Only Africa region is increasing constantly, but other regions are decreasing after 1980.

### Q5
```{r}
gpminder %>% 
  group_by(time, region) %>% 
  summarize(co2_pc = mean(co2_PC, na.rm = TRUE)) %>%
  na.omit() %>%
  filter(time %in% c(1960, 2016)) %>%
  ggplot(aes(x = region, y = co2_pc, fill = factor(time))) + 
  geom_col(position = position_dodge(), color = "black") +
  labs(y = "Average CO2 per capita", 
       x = "Continent",
       y = "Year",
       fill="Year") +
  scale_fill_manual(values = c("darkblue", "gold"))
```

### Q6
```{r}
gpminder %>%
  filter(time == 2016) %>%
  group_by(region) %>% 
  slice_max(co2_PC, n = 3)
```


```{r}
gpminder %>%
  filter(time == 2016) %>%
  group_by(region) %>% 
  slice_min(co2_PC, n = 3)
```

# 4. GDP per capita
### Q1
```{r}
ggplot(gpminder %>% 
         filter(time == 1960, !is.na(region)) , aes(x = lifeExpectancy, y = GDP_PC,
                                        size = totalPopulation, 
                                        color = region
                                        )) + geom_point() 
```

There is a positive relationship between the life expectancy and GDP per capita.

### Q2
```{r}
ggplot(gpminder %>% 
         filter(time == 2019, !is.na(region)), aes(x = lifeExpectancy, y = GDP_PC,
                                        size = totalPopulation, 
                                        color = region
                                        )) + geom_point() 
```

### Q3
The average life expectancy and GPD per capita display a positive relationship between each other. From this we could see the life quality of people around the world are getting better and better throught the last 60 years.

### Q4
```{r}
gpminder %>% 
  filter(!is.na(region)) %>% 
  filter(time %in% c(1960, 2019)) %>%
  group_by(region, time) %>% 
  summarize(lifeExpectancy = mean(lifeExpectancy, na.rm = TRUE)) %>%
  spread(key = time, value = lifeExpectancy)
```

The figures met the result that we calculated

### Q5

```{r}
gpminder %>% 
  group_by(time) %>% 
  mutate(lag_lifeExp = lag(lifeExpectancy), 
         LE_growth = (lifeExpectancy - lag_lifeExp)/lag_lifeExp) %>%
  group_by(region) %>% 
  summarize(growth = mean(LE_growth, na.rm = TRUE)) %>%
  arrange(growth) %>% 
  na.omit()
```

Europe has the highest LE growth rate, and Afirca has the lowest LE growth rate.

### Q6
```{r}
gpminder %>% 
  filter(time %in% c(1960, 2019)) %>%
  ggplot(aes(x = GDP_PC, fill = factor(time))) + geom_histogram(col = "white") +
  labs(fill = "Year", xlab="GDP Per Capita")
```

### Q7
```{r}
gpminder %>% 
  filter(!is.na(name)) %>% 
  group_by(time) %>% 
  mutate(rank = rank(-lifeExpectancy)) %>%
  filter(time %in% c(1960, 2019), name == "United States of America") %>% 
  select(time, rank)
```

### Q8
```{r}
gpminder %>% 
  filter(!is.na(name)) %>% 
  group_by(time) %>% 
  mutate(r = rank(-lifeExpectancy),
         relative_rank = r/sum(!is.na(lifeExpectancy))) %>%
  filter(time %in% c(1960, 2019), name =="United States of America") %>% 
  select(time, relative_rank)
```

### It takes me four hours to finish this PS.